### 2022-03-31 paper/localpsf.tex
editing intro
update paper based on Tucker's comments. Also, working on automated sigma_min based on nearest neighbors

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. H-matrix approximations are highly desirable because fast H-matrix methods can add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known, which allows us to construct the H-matrix. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator actions.
\end{abstract}

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
\label{eq:state_pde}
0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$ \cite{LAPLACEAPPROXIMATION}.

@@

Conventionally, PC approximations are applied to vectors using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as a fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. One may perform these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. Thus, it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. 
While not commonly done, in theory linear algebra operations involving convolution-product approximations (transposes of PC approximations) could be done using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET} since these are equivalent to low rank approximation of the pseudo-differential symbol of the operator. We also note other relevant matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} which do not use PC approximation.

@@

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which we do not know, and which may differ from $\Phi$ due to discretization error. Since the functions $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

>>>>

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. H-matrix approximations are highly desirable because tereh are fast H-matrix methods to add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known, which allows us to construct the H-matrix. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator applies.
\end{abstract}

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
\label{eq:state_pde}
0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$ \cite{LAPLACEAPPROXIMATION}.

@@

Conventionally, PC approximations are applied to vectors using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. One may perform these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. Thus, it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. 
While not commonly done, in theory linear algebra operations involving convolution-product approximations (transposes of PC approximations) could be done using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET} since these are equivalent to low rank approximation of the pseudo-differential symbol of the operator. We also note other relevant matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} which do not use PC approximation.

@@

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which we do not know, and which may differ from $\Phi$ due to discretization error. Since the functions in $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.
